A Formal Practical Foundation of Bidirectional Typing for Simple Types

### 2023-06-09 tex/BiSig.tex
[ tex ] Statements of main results

\usepackage{xfrac,xspace,xcolor,subcaption}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[inline]{enumitem} % for environment enumerate*

\usepackage[draft]{fixme}
\usepackage{subfiles}

@@

\FXRegisterAuthor{LT}{LTan}{L-T}
\FXRegisterAuthor{Josh}{Joshan}{Josh}

@@

\title{A Formal Theory of Bidirectional Typing for Simple Types}

@@

\maketitle
\section{Introduction}\label{sec:intro}

@@

\section{A Case Study: Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus}\label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\begin{remark}
  \begin{LTannote}{$\Xi$ is just a number}
    We will just use $\Xi : \mathbb{N}$ for the set of variables.
  \end{LTannote}

\end{remark}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{\Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
    and $\rho : \Sub{\Xi}{\emptyset}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\begin{LTannote}{deriving from annotation}
  deriving bidirectional binding signature from annotation
\end{LTannote}
\begin{theorem}[Signature annotation]

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}\;d$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

@@

\section{Bidirectional Type Synthesis and Checking}

@@

\subsection{Uniqueness of Synthesis}

\subsection{Type Synthesis and Checking}

>>>>

\usepackage{xfrac,xspace,xcolor,subcaption}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[inline]{enumitem} % for environment enumerate*
\setlist[enumerate]{mode=unboxed}

\setlength {\marginparwidth }{2cm}
\usepackage[color=yellow,textsize=small]{todonotes}
\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}

\usepackage{subfiles}

@@

@@

\title{A Formal Practical Foundation of Bidirectional Typing for Simple Types}

@@

\maketitle

\section{Introduction}\label{sec:intro}

@@

Our assumptions and contributions:
\begin{enumerate}
  \item Syntax-directed typing rules
  \item Simple types
  \item Fully formalised in \Agda 
  \item Generic.
\end{enumerate}

\section{Using Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus as an Outline} \label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{{\color{red}\rho : \Sub{\Xi}{\emptyset}} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\LT{deriving bidirectional binding signature from annotation}

\begin{theorem}[Signature annotation]

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})^\isDir{\chk}}\;(\text{sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

@@

\begin{theorem}[Decidability for Bidirectionalisation]
  For every annotatable raw $t$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
\end{theorem}

\section{Bidirectional Type Synthesis}

@@

\subsection{Decidability of Type Synthesis}

\begin{theorem}[Decidability for Bidirectional Typing Judgements]
  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following two statements hold constructively.
  \begin{enumerate}
    \item For every context $\Gamma$ and a pre-synthesised term $\isTerm{t}$, it is decidable that there exists a type $A$ and a bidirectional typing derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
      \]
    \item For every context $\Gamma$, a pre-checked term $\isTerm{t}$, and a type $A$, it is decidable that there is a typing derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
      \]
  \end{enumerate}
\end{theorem}

\begin{theorem}[Decidability of Type Inference]
  \LT{combine the above two steps}
  Let $\Sigma$ be a type signature and $\Omega$ a binding signature with a bidirectional annotation $\Omega^{\leftrightarrows}$.
  Then, for every context $\Gamma$ and raw term $t$, one and only one of the following statements holds:
  \begin{enumerate}
    \item $t$ has typing derivation $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $\vdash_{\Sigma} A$.
    \item $t$ is not typable under $\Gamma$.
    \item there exists $t'$ such that $t$ needs more type annotations marked in $t'$ to be inferred.
  \end{enumerate}
\end{theorem}

\subsection{Decidability of Bidirectional Type Synthesis and Checking}

\begin{theorem}{Uniqueness of Synthesised Types}\label{thm:unique-syn}

\begin{theorem}
  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following statements hold constructively.
  \begin{enumerate}
    \item For every context $\Gamma$ and a raw term $t^d$ with a mode $d$, 
      \begin{enumerate}
        \item if $d$ is $\Rightarrow$, then it is decidable that there exists a type $A$ and a typing derivation of
          \[\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A;
          \]
        \item if $d$ is $\Leftarrow$, then for every type $A$ it is decidable that there is a typing derivation of
          \[
            \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
          \]
      \end{enumerate}

    \item For every operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$, a context $\Gamma$, and a raw term
      \[
        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
      \]
      that has a mode $d$, it is decidable that there exists a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ and a typing derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
      \]
    \item For every set $\Xi$ of type variables, a list
      \[
        \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_n, d_n, A_n\right)\right]
      \]
      of triples in $\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)$, a \emph{partial substitution} $\rho\colon\PSub{\Xi}{\emptyset}$, a context $\Gamma$, and a list of raw terms $t_i$'s with mode $d_i$ for $i = 1, \ldots, n$, there is either a minimal extension $\rho'$ of $\rho$ such that there is typing derivations of
      \[
        \Gamma, \simsub{\Delta_i}{\rho'} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\rho'}^{d_i}
      \]
      for each $i$, or there is no such extension $\rho'$ such that each raw term $t_i^{d_i}$ has a typing derivation.

    \item For every set $\Xi$ of type variables, a list $\Delta$ of types with variables in $\Xi$, a partial substitution $\rho\colon \PSub{\Xi, \emptyset}$ defined for all types in $\Delta$, and a raw term with mode $d$ then 
      \begin{enumerate}
        \item if $d$ is $\Rightarrow$, then it is decidable that there exists a type $A$ and a typing derivation of
          \[
            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Rightarrow A 
          \]
        \item if $d$ is $\Leftarrow$, then for every type $A$ it is decidable that there is a typing derivation of 
          \[
            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Leftarrow A.
          \]
      \end{enumerate}
  \end{enumerate}
\end{theorem}
By mutual induction on raw terms $t$ with a mode. 
Moreover, if the type $A$ of $t$ is synthesised but checked against $B$, i.e.\ $t\uparrow$, but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
\begin{proof}
\end{proof}
